Measuring contrast processing in the visual system using SSVEP (and VEP?)

### 2024-03-19 review.qmd
Fix issue with ampersands

Neurons in the visual areas of the brain are primarily responsive to differences between light and dark in an image. This property, referred to as contrast, sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity (Campbell & Green, 1965), defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency (Regan, 1966). For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal. The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker produces a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Variants involving square wave flicker can additionally produce more complex spectral components. The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system (Regan & Regan, 1988). SSVEP signals can also be elicited by periodic changes of stimulus properties other than contrast, such as chromaticity, motion, stereo depth, and facial identity or expression (see Norcia et al., 2015, for an overview); however our focus here is on the contrast response.

@@

Contrast Sensitivity: The ability to detect differences in luminance
between an object and its background (Campbell & Green, 1965).

@@

Campbell, F. W., & Green, D. G. (1965). Optical and retinal factors
affecting visual resolution. J Physiology, 181, 576-593.

@@

You can also use this for 'difficult' populations like babies. One
interesting story was about how SSVEP become a replacement for
preferential looking (which was the other way of looking at infant
visual development). See e.g. Davida Teller. SSVEPs allowed people to
make objective measurements of contrast sensitivity development and
deduce that the visual system was more mature (e.g. more functional) in
infancy than previously expected. Also measures of colour sensitivity.
Tinyeyes is based off those measurements. Other people: Tyler, Norcia,
Gunilla H-P, many of the people at SKERI in the 1980s and 90s. Norcia
86,88,90 - mentioned in Regan's nice autobiography :
https://www.google.co.uk/books/edition/Seeing_Spatial_Form/HqaKlR_GpP0C?hl=en&gbpv=1&dq=Analysis+of+EEG+responses+in+man,+evoked+by+sine+watre+modulated+light.+Thesis&pg=PA215&printsec=frontcover

@@

Di Russo, F., Pitzalis, S., Spitoni, G., Aprile, T., Patria, F., Stella,
A., \... & Hillyard, S. A. (2007). Identification of the neural sources
of the pattern-reversal VEP. Neuroimage, 34(1), 177-189.

@@

\[The story goes: Up to the lte 90s people were primarily interested in
measuring contrast sensitivity - the shape of the response function was
assumed to be basically log-linear - and they fit it with straight lines
to extrapolate back to zero response. But then (once Heeger's 1992 paper
had sunk in - see also stuff like Shapley and Victor 1981 [(Shapley &
Victor, 1981)](https://www.zotero.org/google-docs/?zAmtgj)), people
started thinking about gain control - Candy and Norcia in about 1999,
Porciatti, probably a load of Tyler papers that I don't even know
about\...\]. Probably look in p

@@

Ross, J., Speed, H. D., & Morgan, M. J. (1989). The effects of
adaptation and masking on incremental thresholds for contrast. Vision
research, 29(2), 205-215.

Haynes, J. D., Roth, G., Stadler, M., & Heinze, H. J. (2003).
Neuromagnetic correlates of perceived contrast in primary visual cortex.
Journal of Neurophysiology, 89(6), 2655-2666.

Müller, M. M., Picton, T. W., Valdes-Sosa, P., Riera, J.,
Teder-Sälejärvi, W. A., & Hillyard, S. A. (2006). Effects of spatial
selective attention on the steady-state visual evoked potential in the
20--28 Hz range. Cognitive Brain Research, 24(1), 1-13.

@@

Norcia, A. M., Appelbaum, L. G., Ales, J. M., Cottereau, B. R., &
Rossion, B. (2015). The steady-state visual evoked potential in vision
research: A review. Journal of vision, 15(6), 4-4.

@@

Zhu, D., Bieger, J., Molina, G. G., & Aarts, R. M. (2010). A survey of
stimulation methods used in SSVEP-based BCIs. Computational intelligence
and neuroscience, 2010.

>>>>

Neurons in the visual areas of the brain are primarily responsive to differences between light and dark in an image. This property, referred to as contrast, sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity (Campbell \& Green, 1965), defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency (Regan, 1966). For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal. The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker produces a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Variants involving square wave flicker can additionally produce more complex spectral components. The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system (Regan \& Regan, 1988). SSVEP signals can also be elicited by periodic changes of stimulus properties other than contrast, such as chromaticity, motion, stereo depth, and facial identity or expression (see Norcia et al., 2015, for an overview); however our focus here is on the contrast response.

@@

Contrast Sensitivity: The ability to detect differences in luminance
between an object and its background (Campbell \& Green, 1965).

@@

Campbell, F. W., \& Green, D. G. (1965). Optical and retinal factors
affecting visual resolution. J Physiology, 181, 576-593.

@@

You can also use this for 'difficult' populations like babies. One
interesting story was about how SSVEP become a replacement for
preferential looking (which was the other way of looking at infant
visual development). See e.g. Davida Teller. SSVEPs allowed people to
make objective measurements of contrast sensitivity development and
deduce that the visual system was more mature (e.g. more functional) in
infancy than previously expected. Also measures of colour sensitivity.
Tinyeyes is based off those measurements. Other people: Tyler, Norcia,
Gunilla H-P, many of the people at SKERI in the 1980s and 90s. Norcia
86,88,90 - mentioned in Regan's nice autobiography :

@@

Di Russo, F., Pitzalis, S., Spitoni, G., Aprile, T., Patria, F., Stella,
A., \... \& Hillyard, S. A. (2007). Identification of the neural sources
of the pattern-reversal VEP. Neuroimage, 34(1), 177-189.

@@

\[The story goes: Up to the lte 90s people were primarily interested in
measuring contrast sensitivity - the shape of the response function was
assumed to be basically log-linear - and they fit it with straight lines
to extrapolate back to zero response. But then (once Heeger's 1992 paper
had sunk in - see also stuff like Shapley and Victor 1981 [(Shapley \&
Victor, 1981)](https://www.zotero.org/google-docs/?zAmtgj)), people
started thinking about gain control - Candy and Norcia in about 1999,
Porciatti, probably a load of Tyler papers that I don't even know
about\...\]. Probably look in p

@@

Ross, J., Speed, H. D., \& Morgan, M. J. (1989). The effects of
adaptation and masking on incremental thresholds for contrast. Vision
research, 29(2), 205-215.

Haynes, J. D., Roth, G., Stadler, M., \& Heinze, H. J. (2003).
Neuromagnetic correlates of perceived contrast in primary visual cortex.
Journal of Neurophysiology, 89(6), 2655-2666.

Müller, M. M., Picton, T. W., Valdes-Sosa, P., Riera, J.,
Teder-Sälejärvi, W. A., \& Hillyard, S. A. (2006). Effects of spatial
selective attention on the steady-state visual evoked potential in the
20--28 Hz range. Cognitive Brain Research, 24(1), 1-13.

@@

Norcia, A. M., Appelbaum, L. G., Ales, J. M., Cottereau, B. R., \&
Rossion, B. (2015). The steady-state visual evoked potential in vision
research: A review. Journal of vision, 15(6), 4-4.

@@

Zhu, D., Bieger, J., Molina, G. G., \& Aarts, R. M. (2010). A survey of
stimulation methods used in SSVEP-based BCIs. Computational intelligence
and neuroscience, 2010.
